---
title: "Collecting the Cluster Based Tree and Metadata with huebreaker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The goal of huebreaker is to find the cluster based tree and metadata of the data. The package is based on the paper:

Ragonnet-Cronin, M., Hodcroft, E., Hué, S. et al. Automated analysis of phylogenetic clusters. BMC Bioinformatics 14, 317 (2013). <https://doi.org/10.1186/1471-2105-14-317>

Hall M, Woolhouse M, Rambaut A (2015) Epidemic Reconstruction in a Phylogenetics Framework: Transmission Trees as Partitions of the Node Set. PLoS Comput Biol 11(12): e1004613. <https://doi.org/10.1371/journal.pcbi.1004613>

This package is the part of Dhihram Tenrisau, MSc Health Data Science summer project, 'Phylodynamic of Norovirus in UK 2003-2023'. The project is supervised by Stéphane Hué

## Installation

You can install the development version of huebreaker from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Dhihram/huebreaker")
```

## Example

This package needs the additional package `tidyverse`, `ape`, `treeio`, and `dplyr`

```{r, warning=FALSE, message=FALSE}
library(huebreaker)
library(tidyverse)
library(ape)
library(dplyr)
library(treeio)
```

### Data

You need the 2 data in this file, the first data is tree files (newick or nexus) and metadata file. The metadata file consists of `label`, `location`, and `date` columns. The `label` column must be same with the label in tree file.

```{r}
#open data
tree <- read.tree("https://raw.githubusercontent.com/Dhihram/huebreaker/master/data/random_tree.nwk")
metat <- read.csv("https://raw.githubusercontent.com/Dhihram/huebreaker/master/data/random_metadata.csv")

#change date format
metat$date <- as.Date(metat$date, format = "%m/%d/%Y")

#check the data
str(tree)
str(metat)
```

The intial tree can be seen below with `ggtree` package

```{r, message = FALSE, warning = FALSE}
library(ggtree)
gg <- ggtree(tree) + geom_tiplab(size = 2)+
  geom_text2(aes(subset=!isTip, label=label),
             size = 3,
             color = "#0063B1",
             hjust = 1,
             vjust = -1.5) + ggtitle("Random Phylogenetic Tree with Bootstrap Values")
gg
```

### Package Utilization

This package will utilize:
1. Finding and clustering the monophylectic groups in the tree
2. Add the parameter of the clusters: `bootstrap_treshold`, `data_range`, and `samearea`
3. Keep the maximum monophylectic groups in the cluster identify

the `bootstrap_treshold` is the minimum bootstrap value to be considered as a cluster. The `data_range` is the range of the days to be considered as a cluster. The `samearea` is the boolean value to consider the same area as a cluster.

```{r}
res <- huebreaker(tree, metat, bootstrap_threshold = 80, date_range = 30, samearea = TRUE)
knitr::kable(res)
```

We can check the clusters in tree

```{r}
# Highlight all parent nodes
for (node in res$ParentNode) {
  gg_clus <- gg +
    geom_hilight(node = res$ParentNode, fill = "gold", alpha = 0.5)
}
gg_clus
```
